---
title: "Case-Study-Google-Analytics-1 Markdown"
author: "Gado Shekwoguyilo Joshua"
date: "2023-09-03"
output: html_document
---

# Cyclistic Exercise Full Year Analysis

## Analysis on 2019 data to observe difference between usertpes

------------------------------------------------------------------------

## Scope Of The Analysis

This analysis is for case study 1 from the Google Data Analytics Certificate (Cyclistic). It's originally based on the case study "'Sophisticated, Clear, and Polished': Divvy and Data Visualization" written by Kevin Hartman [found here](https://artscience.blog/home/divvy-dataviz-case-study). Our main objective is to explore how annual members and casual riders use Cyclistic bikes differently.

We will be using the Divvy dataset for the case study. The purpose of this script is to consolidate downloaded Divvy data into a single dataframe and then conduct simple analysis to help answer the key question: "In what ways do members and casual riders use Divvy bikes differently?"

*Required packages* The following packages will be required for the analysis : 1. tidyverse for data import and wrangling 2. lubridate for date functions 3. ggplot for visualization

```{r}
install.packages("tidyverse")
install.packages("lubridate")
install.packages("ggplot2")
```

*load packages*

```{r}
library(tidyverse)  ##helps wrangle data
library(lubridate)  ##helps wrangle date attributes
library(ggplot2)  ##helps visualize data
```

### Data Overview

We will download the data for the four quaters of the divvy trips 2019 data from <https://divvy-tripdata.s3.amazonaws.com/index.html>

```{r}
##import divvy_trip_2019 data, Q1-Q4
q1_2019 <- read_csv("Divvy_Trips_2019_Q1.csv")
q2_2019 <- read_csv("Divvy_Trips_2019_Q2.csv")
q3_2019 <- read_csv("Divvy_Trips_2019_Q3.csv")
q4_2019 <- read_csv("Divvy_Trips_2019_Q4.csv")
```

### Data Inspection & Manipulation

we will Check for column name consistency in all csv files then wrangle the data and combine into a single file

```{r}
colnames(q1_2019)
colnames(q2_2019)
colnames(q3_2019)
colnames(q4_2019)
```

Due to a naming difference in q2_2019 columns we will rename them to make it consistent with the rest of the data

```{r}
##To rename q2_2019 columns
q2_2019 <- rename(q2_2019, 
                  trip_id = "01 - Rental Details Rental ID",
                  start_time = "01 - Rental Details Local Start Time",
                  end_time = "01 - Rental Details Local End Time",
                  bikeid = "01 - Rental Details Bike ID",
                  tripduration = "01 - Rental Details Duration In Seconds Uncapped",
                  from_station_id = "03 - Rental Start Station ID",
                  from_station_name = "03 - Rental Start Station Name", 
                  to_station_id = "02 - Rental End Station ID",
                  to_station_name = "02 - Rental End Station Name",
                  usertype = "User Type",
                  gender = "Member Gender",
                  birthyear = "05 - Member Details Member Birthday Year")

##Check all column names again
colnames(q1_2019)
colnames(q2_2019)
colnames(q3_2019)
colnames(q4_2019)
```

Next we inspect the data for Incongruencies, then convert the ride_id and bikeid columns to characters types using as.character

```{r}
# Inspect the dataframes and look for Incongruencies
str(q1_2019)
str(q2_2019)
str(q3_2019)
str(q4_2019)


# Convert trip_id and bikeid to character so that they can stack correctly
q1_2019 <- mutate(q1_2019, trip_id = as.character(trip_id)
                  , bikeid = as.character(bikeid))
q2_2019 <- mutate(q2_2019, trip_id = as.character(trip_id)
                   , bikeid = as.character(bikeid))
q3_2019 <- mutate(q3_2019, trip_id = as.character(trip_id)
                   , bikeid = as.character(bikeid))
q4_2019 <- mutate(q4_2019, trip_id = as.character(trip_id)
                   , bikeid = as.character(bikeid))
```

Now we combine all individual dataframes into one and call it **all_trips**

```{r}
# to combine all data frames into one and call it all_trips
all_trips <- bind_rows(q1_2019, q2_2019, q3_2019, q4_2019)

# To remove birthyear, gender and tripduration fields as this data was dropped beginning in future observations
all_trips <- all_trips %>%  
  select(-c(birthyear, gender, tripduration))
```

### Cleaning up the Data for Analysis

We will inspect the new dataframe **all_trips** that has been created

```{r}
colnames(all_trips)  #List of column names
nrow(all_trips)  #to check how many rows are in data frame
dim(all_trips)  #to check the dimension of the data
head(all_trips)  #See the first 6 rows of data frame
str(all_trips)  #See list of columns and data types (numeric, character, etc)
summary(all_trips)  #Statistical summary of data. Mainly for numerics

```

There are a few problems we will need to fix:

1.  In the "usertype" column, there are two names for members i will be changing the Subscriber and Customer to Member and Casual (as these were the variables stated in the business task.)

2.  The data can only be aggregated at the ride-level, which is too granular. We will want to add some additional columns of data -- such as day, month, year -- that provide additional opportunities to aggregate the data.

3.  We will want to add a calculated field for length of ride since the 2020Q1 data did not have the "tripduration" column. We will add "ride_length" to the entire dataframe for consistency.

4.  There are some rides where tripduration shows up as negative, including several hundred rides where Divvy took bikes out of circulation for Quality Control reasons. We will want to delete these rides.

```{r}
# Begin by seeing how many observations fall under each usertype
table(all_trips$usertype)

# it shows that there are   880637-Customer, 2937367-Subscriber 

# Reassign to the desired values (we will go with "mmeber" and "casual")
all_trips <-  all_trips %>% 
  mutate(usertype = recode(usertype
                                ,"Subscriber" = "member"
                                ,"Customer" = "casual"))
# Check to make sure the proper number of observations were reassigned
table(all_trips$usertype)

```

We will add columns that list the date, month, day, and year of each ride This will allow us to aggregate ride data for each month, day, or year ... before completing these operations we could only aggregate at the ride level

```{r}
all_trips$date <- as.Date(all_trips$start_time) #The default format is yyyy-mm-dd
all_trips$month <- format(as.Date(all_trips$date), "%m")
all_trips$day <- format(as.Date(all_trips$date), "%d")
all_trips$year <- format(as.Date(all_trips$date), "%Y")
all_trips$day_of_week <- format(as.Date(all_trips$date), "%A")

```

We will now add a new column called ride_length, which is a calculation of all trips (in seconds)

```{r}
all_trips$ride_length <- difftime(all_trips$end_time,all_trips$start_time)
# Inspect the structure of the columns
str(all_trips)
#convert ride_length to numeric
all_trips$ride_length <- as.numeric(as.character(all_trips$ride_length))

is.numeric(all_trips$ride_length)
```

To remove Bad data The dataframe includes a few hundred entries when bikes were taken out of docks and checked for quality by Divvy or ride_length was negative We will create a new version of the dataframe (v2) since data is being removed

```{r}
all_trips_v2 <- all_trips[!(all_trips$from_station_name == "HQ QR" | all_trips$ride_length<0),]

```

### Data Analysis

Here we will explore various forms of analysis on the ride_length column (N.B all figures are in seconds), such as mean, median e.t.c

```{r}
mean(all_trips_v2$ride_length) #straight average (total ride length / rides)
median(all_trips_v2$ride_length) #midpoint number in the ascending array of ride lengths
max(all_trips_v2$ride_length) #longest ride
min(all_trips_v2$ride_length) #shortest ride
```

-   we get the mean as; 24.17443
-   we get the median as; 11.81667
-   the max ride_length is; 177200.4
-   the min ride_length is; 1.016667 

 All of this information can also be gotten with the summary() function

```{r}
summary(all_trips_v2$ride_length)
```

To perform descriptive analysis for the two groups in usertype

```{r}
aggregate(all_trips_v2$ride_length ~ all_trips_v2$usertype, FUN = mean)
aggregate(all_trips_v2$ride_length ~ all_trips_v2$usertype, FUN = median)
aggregate(all_trips_v2$ride_length ~ all_trips_v2$usertype, FUN = max)
aggregate(all_trips_v2$ride_length ~ all_trips_v2$usertype, FUN = min)
```

The table below gives the summary of the descriptive analysis done on the usertype groups ( member and casual)

| Ananlysis/Usertype | casual   | member   |
|--------------------|----------|----------|
| Mean               | 57.01802 | 14.32780 |
| Median             | 25.83333 | 9.80000  |
| Max                | 177200.4 | 150943.9 |
| Min                | 1.016667 | 1.016667 |

Next we check for the average ride time for each day of the week for member and causal

```{r}
aggregate(all_trips_v2$ride_length ~ all_trips_v2$usertype + all_trips_v2$day_of_week, FUN = mean)

# to fix the order of the day of week
all_trips_v2$day_of_week <- ordered(all_trips_v2$day_of_week, levels=c("Sunday", "Monday", "Tuesday", "Wednesday", "Thursday", "Friday", "Saturday"))

# Now, let's run the average ride time by each day for members vs casual users
aggregate(all_trips_v2$ride_length ~ all_trips_v2$usertype + all_trips_v2$day_of_week, FUN = mean)

```

to analyze the data by rider type and weekday

```{r}
# analyze ridership data by type and weekday
all_trips_v2 %>% 
  mutate(weekday = wday(start_time, label = TRUE)) %>%  #creates weekday field using wday()
  group_by(usertype, weekday) %>%  #groups by usertype and weekday
  summarise(number_of_rides = n()							#calculates the number of rides and average duration 
            ,average_duration = mean(ride_length)) %>% 		# calculates the average duration
  arrange(usertype, weekday)								# sorts

```

Now we can visualize the data Let's visualize the number of rides by rider type

```{r}
all_trips_v2 %>% 
  mutate(weekday = wday(start_time, label = TRUE)) %>% 
  group_by(usertype, weekday) %>% 
  summarise(number_of_rides = n()
            ,average_duration = mean(ride_length)) %>% 
  arrange(usertype, weekday)  %>% 
  ggplot(aes(x = weekday, y = number_of_rides, fill = usertype)) +
  geom_col(position = "dodge")
```

Let's create a visualization for average duration

```{r}
all_trips_v2 %>% 
  mutate(weekday = wday(start_time, label = TRUE)) %>% 
  group_by(usertype, weekday) %>% 
  summarise(number_of_rides = n()
            ,average_duration = mean(ride_length)) %>% 
  arrange(usertype, weekday)  %>% 
  ggplot(aes(x = weekday, y = average_duration, fill = usertype)) +
  geom_col(position = "dodge")
```

